Holomorphic field theories and higher algebra

### 2025-01-29 draft.tex
another pass

Before saying anything technical, we indicate how derived geometry generalizes more familiar bits of mathematics.
The starting place is the theory of vector spaces, which is very concrete and provides a rich source of both geometric intuitions and algebraic tools.
This linear setting generalizes in two directions:
\begin{itemize}
\item the geometry of nonlinear spaces, whether in the style of algebraic geometry or in the style of differential geometry, and
\item homological algebra, or differential graded vector spaces (and modules, sheaves, and so on).
\end{itemize} 
Both of these generalizations pass back to linear algebra in natural ways:
\begin{itemize}
\item at a point on a manifold or variety, the tangent space offers the best linear approximation to the space at that point (where it makes sense), and
\item given a cochain complex $(V,d)$, any cohomology group $H_k(V)$ is a vector space encoding essential information about the complex.
\end{itemize} 
As a first pass at derived geometry, one might say that it seeks a common generalization of nonlinear geometry and homological algebra.
With slightly more precision, one might ask for a notion of space such that each point has a tangent {\em complex}, i.e., the linearization of a point yields a cochain complex.

Note that there will be different versions of derived geometry, depending on the kind of nonlinear geometry concerned: 
there is derived algebraic geometry, derived differential geometry, derived symplectic geometry, and so on.
A desideratum of derived geometry is then that 
\begin{itemize}
\item an ordinary scheme or manifold (or whatever class of underived space) should provide an example of a derived space, and
\item any cochain complex should provide an example of a derived space.
\end{itemize}
That is, derived geometry contains both the domains it seeks to generalize.

@@

Derived stacks then arise from derived schemes by intelligently taking quotients and, more generally, colimits.
Recall that a scheme in algebraic geometry has a functor of points valued in $\Sets$ while a stack has a functor of points valued in $\Gpds$, 
the category of groupoids.\footnote{Every equivalence relation determines a groupoid where the relation gives the morphisms. This groupoid knows a lot more than the underlying quotient set, which is known as the ``connected components'' of this groupoid. In this way, groupoids offer a refinement of traditional quotients.}
More accurately we work with groupoids localized at Morita equivalence ({\it cf.} quasi-isomorphism of cochain complexes and the derived category).
In derived geometry we push farther and work with topological spaces up to weak homotopy equivalences (or simplicial sets up to weak homotopy equivalence),
which remembers more;
every topological space has an underlying fundamental groupoid.
Thus the nature of derived geometry is indicated by the (co)domains of the functors of points:
\[
\begin{tikzcd}[column sep=large]
\CAlg^{\dg}_{\leq 0} \arrow[dd, bend right=40, "H^0"'] \arrow[rr, "\text{derived stack}"] & & \Spaces \arrow[d, bend left=40, "\Pi_{\leq 1}"]\\
&& \Gpds \arrow[d, bend left=40,  "\pi_0"] \arrow[u, hook]\\
\CAlg \arrow[uu, hook] \arrow[rr, "\text{scheme}" near end] \arrow[urr, "\text{stack}"] && \Sets \arrow[u, hook]
\end{tikzcd}
\]
(This diagram is purloined from the wonderful (and short!) ``What is derived algebraic geometry?''~\cite{VezAMS}.)
A derived stack admits a truncation to an ordinary stack by restricting the domain category and mapping down to groupoids as the codomain category.

To ``derive'' a notion from traditional geometry, one typically needs to recast it using the functor of points and then ponder how its features might generalize to derived affine schemes and~$\Spaces$.
By now many such notions have received such a treatment, including analytic geometry~\cite{} and Poisson geometry~\cite{}.

@@

The basic ingredients of classical field theory are a ``spacetime'' manifold $M$, the ``fields'' given by sections of some fiber bundle $E \to M$, and a system of PDE on the fields known as the ``equations of motion.''
The essential goal is to describe the space of solutions to these equations of motion.
All these features admit derived versions: one can replace the manifold $M$ by a derived space $\MM$ of some kind and the fiber bundle by a map of derived spaces $\EE \to \MM$ and the fields by sections thereof.
The theory of differential equations admits a generalization too (perhaps most cleanly by using the perspective of $D$-modules and its outgrowths).
In practice, as we'll see in this paper, many of the moduli spaces are variants of spaces already studied by geometers.
In fact, they are often ``derived enhancements'' of well-known moduli spaces: we work with a derived stack whose truncation is a stack already known.
For example, holomorphic Chern-Simons theory is best viewed as studying the derived enhancement of the moduli space of $G$-bundles on a Calabi-Yau 3-fold~$M$, where $G$ is a reductive complex algebraic group.

There is an important feature that we have not discussed so far:
the equations of motion arise from ``variational calculus.''
Loosely speaking, these equations identify the critical points of some function on the space of fields.
When approached from a derived perspective, 
the space of solutions typically obtains a natural $-1$-shifted symplectic structure.\footnote{The derived critical locus of a function $f$ on an ordinary smooth variety $X$ is the derived intersection inside $T^*X$ of the graph of $\d f$ and the zero section. The structure sheaf can be modeled by functions $\cO(T^*[-1]X)$ equipped with a differential determined by~$f$. The shifted symplectic structure on $T^*[-1]X$ carries over to this derived critical locus.}
Hence derived field theories lead to $-1$-shifted symplectic derived stacks,
and we have examples like $T^*[-1]\Bun_G(M)$ or~$T^*[-1] \RR\Map(M,X)$.

>>>>

Before saying anything technical, we indicate how derived geometry generalizes more familiar bits of mathematics.
The starting place is the theory of vector spaces, which is very concrete and provides a rich source of both geometric intuitions and algebraic tools.
This linear setting generalizes in two directions:
\begin{itemize}
\item the geometry of nonlinear spaces, whether in the style of algebraic geometry or in the style of differential geometry, and
\item homological algebra, or differential graded vector spaces (and modules, sheaves, and so on).
\end{itemize} 
Both of these generalizations pass back to linear algebra in natural ways:
\begin{itemize}
\item at a point $x$ on a manifold $M$ (or variety), the tangent space $T_xM$ offers the best linear approximation to the space at that point (where it makes sense), and
\item given a cochain complex $(V,d)$, a cohomology group $H^0(V)$ is a vector space encoding essential information about the complex.
\end{itemize} 
As a first pass at derived geometry, one might say that it seeks a common generalization of nonlinear geometry and homological algebra.
With slightly more precision, one might ask for a notion of derived space $M$ such that
\begin{itemize} 
\item each point $x$ in $M$ has a tangent {\em complex} $\TT_x M$, i.e., the linearization of a point yields a cochain complex, and
\item $M$ should have a truncation $\tau(M)$ to an ordinary space.
\end{itemize}
Note that there will be different versions of derived geometry, depending on the kind of nonlinear geometry concerned: 
there is derived algebraic geometry, derived differential geometry, derived symplectic geometry, and so on.

\begin{center}
\begin{tikzcd}
&\text{linear algebra} \arrow[squiggly, dr] \arrow[squiggly, dl] &\\
\text{homological algebra} \arrow[squiggly, dr] \arrow[ur, bend left, "H^0"] &&\text{nonlinear geometry}\arrow[squiggly, dl] \arrow[ul, bend right, "\rm{T}"']\\
&\text{derived geometry} \arrow[ul, bend left, "\TT"] \arrow[ur, bend right, "\tau"']&
\end{tikzcd}
\end{center}

A related desideratum of derived geometry is that 
\begin{itemize}
\item an ordinary scheme or manifold (or whatever class of underived space) should provide an example of a derived space, and
\item any cochain complex should provide an example of a derived space.
\end{itemize}
That is, derived geometry contains both the domains it seeks to generalize.

@@

Derived stacks then arise from derived schemes by intelligently taking quotients and, more generally, colimits.
Recall that a scheme in algebraic geometry has a functor of points valued in $\Sets$ while a stack has a functor of points valued in $\Gpds$, 
the category of groupoids.\footnote{Every equivalence relation determines a groupoid where the relation gives the morphisms. This groupoid knows a lot more than the underlying quotient set, which is known as the ``connected components'' of this groupoid. In this way, groupoids offer a refinement of traditional quotients.}
More accurately we work with groupoids localized at Morita equivalence ({\it cf.} quasi-isomorphism of cochain complexes and the derived category).
In derived geometry we push farther and work with topological spaces up to weak homotopy equivalences (or simplicial sets up to weak homotopy equivalence),
which remembers more;
every topological space has an underlying fundamental groupoid.
Thus the nature of derived geometry is indicated by the (co)domains of the functors of points:
\[
\begin{tikzcd}[column sep=large]
\CAlg^{\dg}_{\leq 0} \arrow[dd, bend right=40, "H^0"'] \arrow[rr, "\text{derived stack}"] & & \Spaces \arrow[d, bend left=40, "\Pi_{\leq 1}"]\\
&& \Gpds \arrow[d, bend left=40,  "\pi_0"] \arrow[u, hook]\\
\CAlg \arrow[uu, hook] \arrow[rr, "\text{scheme}" near end] \arrow[urr, "\text{stack}"] && \Sets \arrow[u, hook]
\end{tikzcd}
\]
(This diagram is purloined from Vezzosi's wonderful (and short!) ``What is \dots a derived stack?''~\cite{VezAMS}.)
A derived stack admits a truncation to an ordinary stack by restricting the domain category and mapping down to groupoids as the codomain category.

To ``derive'' a notion from traditional geometry, one typically needs to recast it using the functor of points and then ponder how its features might generalize to derived affine schemes and~$\Spaces$.
By now many such notions have received such a treatment, including analytic geometry~\cite{Porta} and Poisson geometry~\cite{CPTVV}.

@@

The basic ingredients of classical field theory are a ``spacetime'' manifold $M$, the ``fields'' given by sections of some fiber bundle $E \to M$, and a system of PDE on the fields known as the ``equations of motion.''
The essential goal is to describe the space of solutions to these equations of motion.
All these features admit derived versions: one can replace the manifold $M$ by a derived space $\MM$ of some kind and the fiber bundle by a map of derived spaces $\EE \to \MM$ and the fields by sections thereof.
The theory of differential equations admits a generalization too (perhaps most cleanly by using the perspective of $D$-modules and its outgrowths).
In practice, as we'll see in this paper, many of the moduli spaces are variants of spaces already studied by geometers.
In fact, they are often ``derived enhancements'' of well-known moduli spaces: we work with a derived stack whose truncation is a stack already known.\footnote{In \cite{STV} a number of important moduli spaces, like moduli of stable maps of curves or moduli of perfect complexes, receive derived enhancements.}
For example, holomorphic Chern-Simons theory is best viewed as studying the derived enhancement of the moduli space of $G$-bundles on a Calabi-Yau 3-fold~$M$, where $G$ is a reductive complex algebraic group.

There is an important feature that we have not discussed so far:
the equations of motion arise from ``variational calculus.''
Loosely speaking, these equations identify the critical points of some function on the space of fields.
When approached from a derived perspective, 
the space of solutions typically obtains a natural $-1$-shifted symplectic structure.\footnote{The derived critical locus of a function $f$ on an ordinary smooth variety $X$ is the derived intersection inside $T^*X$ of the graph of $\d f$ and the zero section. The structure sheaf can be modeled by functions $\cO(T^*[-1]X)$ equipped with a differential determined by~$f$. The shifted symplectic structure on $T^*[-1]X$ carries over to this derived critical locus. See \cite{VezdCrit} for more discussion.}
Hence derived field theories lead to $-1$-shifted symplectic derived stacks,
and we have examples like $T^*[-1]\Bun_G(M)$ or~$T^*[-1] \RR\Map(M,X)$.
